Notes on Validated Model Counting \\ Version of \today

### 2022-05-02 doc/mc-notes.tex
Change notation
Incorporate top-down proof example

\newcommand{\turnstile}{\vdash}

@@

@@

A {\em separable schema} is a directed acyclic graph representing a
Boolean formula where the only allowed operations are $\neg$, $\pand$,
$\por$.
%% and $\pite$.
We use a graph representation to allow sharing of
subformulas.  The set of schemas over a set of variables
$\{x_1, x_2, \ldots, x_n\}$
can be defined recursively, as is shown in Table
\ref{tab:schema}.  Each schema $S$ has an associated dependency set
$\depend(S)$ and an associated set of models $\interpset{S}$.
%% We
%% divide the operations into {\em basic} ones that are fundamental to
%% the representation, and {\em composite} ones that can be constructed
%% from basic operations.  The latter category consists of just the $\pite$ operation.

@@

@@

@@

A CRAT proof follows the same general form as a QRAT dual proof~\cite{bryant:cade:2021}---one
that ensures that each clause addition and each clause deletion
preserves equivalence.  Starting with the set of input clauses, it
produces a sequence of steps that both add and delete clauses.  Each
addition must be truth preserving and each deletion must be falsehood
preserving.  At the end, all input clauses must have been deleted, and
among the remaining clauses there must be only a single unit clause
consisting of some variable or its complement.  Except for trivial
cases, the final literal will be an extension variable or its
complement.  That literal will indicate the root of the schema as the
(possibly negated) outpuf of a schema operation.  Working from that
root backward, the schema can be extracted from the CRAT file.

@@

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) Schema   \\[2ex]
    \centering{\input{dd/c3}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Cost Computation\\[2ex]
    \centering{\input{dd/c3-eval}}
    \end{minipage}
  }
\caption{Schema for Formula $\phi_1 = x_1 \lor x_2 \lor x_3$ and its Cost Computation}
\label{fig:c3:schema}
\end{figure}

\begin{figure}
  \centering{
  \begin{tabular}{llllll}
    \multicolumn{4}{c}{Proof line} & & \multicolumn{1}{c}{Explanation} \\
\midrule
    \makebox[5mm][l]{\tt 1} & \makebox[7mm][l]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[15mm]{}          & \makebox[5mm]{} & \makebox[20mm][l]{Input clause}\\
            & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand x_3$ \\
    {\tt 2} & {\tt ab}  & {\tt 4  2 -3 0}  & {\tt 0}    & & Defining clauses for $p_4$ \\ 
    {\tt 3} & {\tt ab}  & {\tt -4 -2 0}    & {\tt -2 0} & & \\ %%Defining clause for $p_4$ \\ 
    {\tt 4} & {\tt ab}  & {\tt -4 3 0}    & {\tt -2 0} & & \\ %%Defining clause for $p_4$ \\ 
            & {\tt s}   & {\tt 5 2 4}   & {\tt 3 0}  & & Declare $s_5 = x_2 \por p_4$ \\
    {\tt 5} & {\tt ab}  & {\tt -5 2 4 0}  & {\tt 0}    & & Defining clauses for $s_5$ \\ 
    {\tt 6} & {\tt ab}  & {\tt  5 -2 0}    & {\tt -5 0} &  \\  %%Defining clause for $s_5$ \\ 
    {\tt 7} & {\tt ab}  & {\tt  5 -4 0}    & {\tt -5 0} & &  \\  %%Defining clause for $s_5$ \\ 
            & {\tt p}   & {\tt 6 -1 5}  &            & & Declare $p_6 = \obar{x}_1 \pand s_5$ \\
    {\tt 8} & {\tt ab}  & {\tt 6  1 -5 0}  & {\tt 0}    & & Defining clauses for $p_6$ \\ 
    {\tt 9} & {\tt ab}  & {\tt -6 -1 0}    & {\tt -8 0} & &  \\  %%Defining clause for $p_6$ \\ 
    {\tt 10} & {\tt ab}  & {\tt -6 5 0}    & {\tt -8 0} & &  \\  %%Defining clause for $p_6$ \\ 
             & {\tt s}   & {\tt 7 1 6}   & {\tt 9 0}  & & Declare $s_7 = x_1 \por p_6$ \\
    {\tt 11} & {\tt ab}  & {\tt -7 1 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
    {\tt 12} & {\tt ab}  & {\tt  7 -1 0}    & {\tt -12 0} & &  \\  %%Defining clause for $s_7$ \\ 
    {\tt 13} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -12 0} & &  \\  %%Defining clause for $s_7$ \\ 
    {\tt 14} & {\tt ar}  & {\tt 7 0} & {\tt 12 13 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
             & {\tt dr}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
  \end{tabular}
  }
  \caption{CRAT Proof Steps for Formula $x_1 \lor x_2 \lor x_3$}
  \label{fig:c3:crat}
\end{figure}

@@

\subsection{Example 2}

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) Schema   \\[2ex]
    \centering{\input{dd/p2}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Cost Computation\\[2ex]
    \centering{\input{dd/p2-eval}}
    \end{minipage}
  }
\caption{Schema for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
\label{fig:p2:schema}
\end{figure}

@@

    {\tt 2} & {\tt i}   & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
            & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
    {\tt 3} & {\tt ab}  & {\tt 4 2 3 0} & {\tt 0}    & & Defining clauses for $p_4$ \\
    {\tt 4} & {\tt ab}  & {\tt -4 -2 0} & {\tt -2 0}    & & \\
    {\tt 5} & {\tt ab}  & {\tt -4 -3 0} & {\tt -2 0}    & & \\
            & {\tt p}   & {\tt 5 -1 4}  &             & & Declare $p_5 = \obar{x}_1 \pand p_4 = \obar{x}_1 \pand \obar{x}_2 \pand \obar{x}_3$ \\
    {\tt 6} & {\tt ab}  & {\tt 5 1 -4 0} & {\tt 0}    & & Defining clauses for $p_5$ \\
    {\tt 7} & {\tt ab}  & {\tt -5 -1 0} & {\tt -6 0}    & & \\
    {\tt 8} & {\tt ab}  & {\tt -5 4 0} & {\tt -6 0}    & & \\
    {\tt 9} & {\tt ar}  & {\tt -5 0} & {\tt 7 8 4 5 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
            & {\tt dr}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\
            & {\tt p}   & {\tt 6 1 -2} &              & & Declare $p_6 = x_1 \pand \obar{x}_2$ \\    
    {\tt 10} & {\tt ab}  & {\tt 6 -1 2 0}  & {\tt 0}    & & Defining clauses for $p_6$ \\ 
    {\tt 11} & {\tt ab}  & {\tt -6 1 0}    & {\tt -10 0} & & \\
    {\tt 12} & {\tt ab}  & {\tt -6 -2 0}    & {\tt -10 0} & & \\ 
    {\tt 13} & {\tt ar} & {\tt -6 0} & {\tt 11 12 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
            & {\tt dr}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
            & {\tt s}   & {\tt 7 5 6}   & {\tt 7 11 0}  & & Declare $s_7 = \obar{t}_1 \por \obar{t}_2$ \\
    {\tt 14} & {\tt ab}  & {\tt -7 5 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
    {\tt 15} & {\tt ab}  & {\tt  7 -5 0}    & {\tt -5 0} &  \\  
    {\tt 16} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -5 0} & &  \\
    {\tt 17} & {\tt ar}  & {\tt -7 0}       & {\tt 9 13 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
             & {\tt dr}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
             & {\tt dr}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
  \end{tabular}
}  
  \caption{CRAT Proof Steps for Formula $\phi = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$}
  \label{fig:p2:crat}
\end{figure}

As a more complex example, consider the Boolean formula given by the
conjunction of clauses $C_1 = x_1 \lor x_2 \lor x_3$ and $C_2 =
\obar{x}_1 \lor x_2$.  With this example, we also demonstrate the use
of DeMorgan's Laws to provide a more compact encoding of the formula.

The complement of the formula can be written in DNF as
$\obar{x}_1\,\obar{x}_2\,\obar{x}_3 \lor x_1\,\obar{x}_2$.  Each of
the two conjuncts can be formed using $\pand$ operations, and their
sum can be formed using a $\por$ operation.  In this example, we first
form a term $t_1$ that equals $C_1$ and a term $t_2$ that equals
$C_2$.  (Note that $t_1$ in this case is logically equivalent to $t_1$
in the example of Figure~\ref{fig:c3:schema}, but the use of
negation enables it to be represented with two operations
rather than four.)
Terms $t_1$ and $t_2$ are asserted as unit clauses in proof lines 9 and 13,
allowing the input clauses to be deleted.  Then we generate $t_3$ as
the conjunction of $t_1$ and $t_2$ and assert it as a unit clause.
Based on this, we can delete the unit clauses for terms $t_1$ and
$t_2$.  Term $t_3$ then becomes the root of the schematic representation shown in Figure~\ref{fig:p2:schema}.

@@

Given an arbitrary CNF formula, we can use BDD operations to generate
a schematic representation.  The proof generation can follow the
methods we have used for generating unsatisfiability proofs of Boolean
formulas~\cite{bryant:tacas:2021} and dual proofs of quantified
Boolean formulas~\cite{bryant:cade:2021}.  The key idea is to use
extended resolution to encode the semantics of the BDD nodes as part
of the proof.  Depending on which, if any, of its children are
terminal nodes, a BDD node will be represented in the schema as either
1) a literal, 2) a single $\pand$ operation, or 3) two $\pand$
operations and one $\por$ operation.

Other model counters use either top-down and bottom-up methods to
generate representations of the formula that are similar to our
schemas.  These exploit the key abstractions we have identified.  We
must find ways to modify these model counters to also generate CRAT
proofs.

>>>>

\newcommand{\turnstile}{\models}

@@

Computing the cost function for a Boolean formula becomes
straightforward when the formula contains only the operations
$\pand$, $\por$, and $\neg$, as is demonstrated by
Propositions~\ref{prop:negation}--\ref{prop:disjunction}.  We define {\em separable schemas}
as a direct-acyclic graph representation of such formulas.
Representing formulas as a graph enables sharing subformulas, yielding a more compact representation.

\subsection{Schema Definition}

@@

The set of schemas over a set of variables
$\{x_1, x_2, \ldots, x_n\}$
can be defined recursively, as is shown in Table
\ref{tab:schema}.  Each schema $S$ has an associated dependency set
$\depend(S)$ and an associated set of models $\interpset{S}$.
%% We
%% divide the operations into {\em basic} ones that are fundamental to
%% the representation, and {\em composite} ones that can be constructed
%% from basic operations.  The latter category consists of just the $\pite$ operation.

@@

\subsection{Schema Normalization}

@@

\subsection{Encoding the ITE Operation}
\label{sec:ite}

The if-then-else (ITE) operation arises when converting the CNF
representation of a formula into a separable schema, both for
bottom-up approaches based on decision diagrams, and for top-down
approaches based on CDCL\@.  For functions $f_1$, $f_2$, and $f_3$, we
define
$\ite(f_1, f_2, f_3) = (f_1 \land f_2) \lor (\neg f_1 \land f_3)$.
Observe that the $\lor$ operation in this expression satisfies
the requirements for $\por$, since the first argument can only yield
$\tautology$ for assignments that yield $\tautology$ for $f_1$, while
the second can only yield $\tautology$ for assignments that yield
$\nil$ for $f_1$.  Therefore, the only condition imposed on an
expansion of $\ite$ into the allowed schema operations is that the
dependency set for $f_1$ must be disjoint from those of $f_2$ and
$f_3$.

\begin{table}
  \caption{Encodings of the ITE Operation}
  \label{tab:ite}
  \begin{center}
  \begin{tabular}{ccc}
    ITE Form & & Encoding \\
    \midrule
     $\ite(S_1, S_2, S_3)$ & $$ & $(S_1 \pand S_2) \por (\neg S_1 \land S_3)$ \\
     $\ite(\tautology, S_2, S_3)$ & $$ & $S_2$ \\
     $\ite(\nil, S_2, S_3)$ & $$ & $S_3$ \\
     $\ite(S_1, \tautology, \nil)$ & $$ & $S_1$ \\
     $\ite(S_1, \nil, \tautology)$ & $$ & $\neg S_1$ \\
     $\ite(S_1, S_2, \nil)$ & $$ & $S_1 \pand S_2$ \\
     $\ite(S_1, \nil, S_3)$ & $$ & $\neg S_1 \pand S_3$ \\
     $\ite(S_1, S_2, \tautology)$ & $$ & $\neg (S_1 \pand \neg S_2)$ \\
     $\ite(S_1, \tautology, S_3)$ & $$ & $\neg (\neg S_1 \pand \neg S_3)$ \\
  \end{tabular}
  \end{center}
\end{table}

Table~\ref{tab:ite} shows different ways to encode an ITE operation
into schema operations.  All of these require the dependency set of
the argument $S_1$ to be disjoint from those of arguments $S_2$ and
$S_3$.  The first row shows the most general case, require one $\por$
and two $\pand$ operations.  The other rows show special cases, where
one or more argument is a constant.  Of these, the final two rows are
particularly noteworthy.  They make use of DeMorgan's Laws to convert
disjunctions into conjunctions.  These conjunctions can then be
encoded with $\pand$ operations, since their arguments will have
disjoint dependency sets.

@@

A CRAT proof follows the same general form as a QRAT dual proof~\cite{bryant:cade:2021}---one
that ensures that each clause addition and each clause deletion
preserves equivalence.  Starting with the set of input clauses, it
produces a sequence of steps that both add and delete clauses.  Each
addition must be truth preserving and each deletion must be falsehood
preserving.  At the end, all input clauses must have been deleted, and
among the remaining clauses there must be only a single unit clause
consisting of some variable or its complement.  Except for trivial
cases, the final literal will be an extension variable or its
complement.  That literal will indicate the root of the schema as the
(possibly negated) output of a schema operation.  Working from that
root backward, the schema can be extracted from the CRAT file.

@@

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) Schema   \\[2ex]
    \centering{\input{dd/c3}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Cost Computation\\[2ex]
    \centering{\input{dd/c3-eval}}
    \end{minipage}
  }
\caption{Schema \#1 for Formula $\phi_1 = x_1 \lor x_2 \lor x_3$ and its Cost Computation}
\label{fig:c3:schema}
\end{figure}

\begin{figure}
  \centering{
  \begin{tabular}{llllll}
    \multicolumn{4}{c}{Proof line} & & \multicolumn{1}{c}{Explanation} \\
\midrule
    \makebox[5mm][l]{\tt 1} & \makebox[7mm][l]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[15mm]{}          & \makebox[5mm]{} & \makebox[20mm][l]{Input clause}\\
            & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand x_3$ \\
    {\tt 2} & {\tt ab}  & {\tt 4  2 -3 0}  & {\tt 0}    & & Defining clauses for $p_4$ \\ 
    {\tt 3} & {\tt ab}  & {\tt -4 -2 0}    & {\tt -2 0} & & \\ %%Defining clause for $p_4$ \\ 
    {\tt 4} & {\tt ab}  & {\tt -4 3 0}    & {\tt -2 0} & & \\ %%Defining clause for $p_4$ \\ 
            & {\tt s}   & {\tt 5 2 4}   & {\tt 3 0}  & & Declare $s_5 = x_2 \por p_4$ \\
    {\tt 5} & {\tt ab}  & {\tt -5 2 4 0}  & {\tt 0}    & & Defining clauses for $s_5$ \\ 
    {\tt 6} & {\tt ab}  & {\tt  5 -2 0}    & {\tt -5 0} &  \\  %%Defining clause for $s_5$ \\ 
    {\tt 7} & {\tt ab}  & {\tt  5 -4 0}    & {\tt -5 0} & &  \\  %%Defining clause for $s_5$ \\ 
            & {\tt p}   & {\tt 6 -1 5}  &            & & Declare $p_6 = \obar{x}_1 \pand s_5$ \\
    {\tt 8} & {\tt ab}  & {\tt 6  1 -5 0}  & {\tt 0}    & & Defining clauses for $p_6$ \\ 
    {\tt 9} & {\tt ab}  & {\tt -6 -1 0}    & {\tt -8 0} & &  \\  %%Defining clause for $p_6$ \\ 
    {\tt 10} & {\tt ab}  & {\tt -6 5 0}    & {\tt -8 0} & &  \\  %%Defining clause for $p_6$ \\ 
             & {\tt s}   & {\tt 7 1 6}   & {\tt 9 0}  & & Declare $s_7 = x_1 \por p_6$ \\
    {\tt 11} & {\tt ab}  & {\tt -7 1 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
    {\tt 12} & {\tt ab}  & {\tt  7 -1 0}    & {\tt -12 0} & &  \\  %%Defining clause for $s_7$ \\ 
    {\tt 13} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -12 0} & &  \\  %%Defining clause for $s_7$ \\ 
    {\tt 14} & {\tt ar}  & {\tt 7 0} & {\tt 12 13 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
             & {\tt dr}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
  \end{tabular}
  }
  \caption{CRAT Proof \#1 for Formula $x_1 \lor x_2 \lor x_3$}
  \label{fig:c3:crat}
\end{figure}

@@

\subsection{Example 2A}

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) Schema   \\[2ex]
    \centering{\input{dd/p2-bdd}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Cost Computation\\[2ex]
    \centering{\input{dd/p2-bdd-eval}}
    \end{minipage}
  }
  \caption{Schema \#2A for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
\label{fig:p2-bdd:schema}
\end{figure}

@@

    {\tt 2} & {\tt i}   & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
            & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
    {\tt 3} & {\tt ab}  & {\tt 4 2 3 0} & {\tt 0}    & & Defining clauses for $p_4$ \\
    {\tt 4} & {\tt ab}  & {\tt -4 -2 0} & {\tt -2 0}    & & \\
    {\tt 5} & {\tt ab}  & {\tt -4 -3 0} & {\tt -2 0}    & & \\
            & {\tt p}   & {\tt 5 -1 4}  &             & & Declare $p_5 = \obar{x}_1 \pand p_4 = \obar{x}_1 \pand \obar{x}_2 \pand \obar{x}_3$ \\
    {\tt 6} & {\tt ab}  & {\tt 5 1 -4 0} & {\tt 0}    & & Defining clauses for $p_5$ \\
    {\tt 7} & {\tt ab}  & {\tt -5 -1 0} & {\tt -6 0}    & & \\
    {\tt 8} & {\tt ab}  & {\tt -5 4 0} & {\tt -6 0}    & & \\
    {\tt 9} & {\tt ar}  & {\tt -5 0} & {\tt 7 8 4 5 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
            & {\tt dr}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\
            & {\tt p}   & {\tt 6 1 -2} &              & & Declare $p_6 = x_1 \pand \obar{x}_2$ \\    
    {\tt 10} & {\tt ab}  & {\tt 6 -1 2 0}  & {\tt 0}    & & Defining clauses for $p_6$ \\ 
    {\tt 11} & {\tt ab}  & {\tt -6 1 0}    & {\tt -10 0} & & \\
    {\tt 12} & {\tt ab}  & {\tt -6 -2 0}    & {\tt -10 0} & & \\ 
    {\tt 13} & {\tt ar} & {\tt -6 0} & {\tt 11 12 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
            & {\tt dr}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
            & {\tt s}   & {\tt 7 5 6}   & {\tt 7 11 0}  & & Declare $s_7 = \obar{t}_1 \por \obar{t}_2$ \\
    {\tt 14} & {\tt ab}  & {\tt -7 5 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
    {\tt 15} & {\tt ab}  & {\tt  7 -5 0}    & {\tt -14 0} &  \\  
    {\tt 16} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -14 0} & &  \\
    {\tt 17} & {\tt ar}  & {\tt -7 0}       & {\tt 9 13 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
             & {\tt dr}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
             & {\tt dr}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
  \end{tabular}
}  
  \caption{CRAT Proof \#2A for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$}
  \label{fig:p2-bdd:crat}
\end{figure}

As a more complex example, consider the Boolean formula $\phi_2$ given
by the conjunction of clauses $C_1 = x_1 \lor x_2 \lor x_3$ and $C_2 =
\obar{x}_1 \lor x_2$.  With this example, we also demonstrate the use
of DeMorgan's Laws to provide a more compact encoding of the formula,
similar to the use of these laws when encoding ITE operations in
Table~\ref{tab:ite}.

Figure~\ref{fig:p2-bdd:schema} shows a schema representing the formula,
and Figure~\ref{fig:p2-bdd:crat} shows the associated CRAT proof.  This
proof was generated via a bottom-up strategy, such as would be created
using BDDs.  It creates schematic representations of the input
clauses and then forms their conjunction.  In the proof, unit
clauses are generated for the representations of the input clauses,
and then the input clauses are deleted.  These intermediate unit
clauses are used to justify a unit clause for the final root, and then
they are deleted.

Using DeMorgan's Laws, $C_1$ can be written as 
$t_1 = \neg [\obar{x}_1\land \obar{x}_2\land \obar{x}_3]$, and this can be
expressed using $\pand$ operations $s_4$ and $s_5$, shown in
Figure~\ref{fig:p2-bdd:schema}A\@.  (Note that $t_1$ in this case is
logically equivalent to root $t_1$ in the schema of
Figure~\ref{fig:c3:schema}A, but the use of negation enables it to be
represented with two operations rather than four.)  Similarly, $C_2$
can be written as $t_2 = \neg [x_1\land \obar{x}_2]$, and this can be
represented by $\pand$ operation $p_6$.
Terms $t_1$ and $t_2$ are asserted as unit clauses on proof lines 9 and 13,
allowing the input clauses to be deleted.
The conjunction $t_1 \land t_2$
can be written as $t_3 = \neg[\obar{t_1} \lor \obar{t_2}]$, represented by $\por$ operation $s_7$.
Term $t_3$ is asserted as a unit clause on proof line 17.
Based on this, the unit clauses for terms $t_1$ and
$t_2$ can be deleted.  Term $t_3$ then becomes the unique root of the cost
computation shown in Figure~\ref{fig:p2-bdd:schema}B\@.

\subsection{Example 2B}

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) Schema   \\[2ex]
    \centering{\input{dd/p2-cdcl}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Cost Computation\\[2ex]
    \centering{\input{dd/p2-cdcl-eval}}
    \end{minipage}
  }
\caption{Schema \#2B for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
\label{fig:p2-cdcl:schema}
\end{figure}

    {\tt 2} & {\tt i}   & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
            & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
    {\tt 3} & {\tt ab}  & {\tt 4 2 3 0} & {\tt 0}    & & Defining clauses for $p_4$ \\
    {\tt 4} & {\tt ab}  & {\tt -4 -2 0} & {\tt -2 0}    & & \\
    {\tt 5} & {\tt ab}  & {\tt -4 -3 0} & {\tt -2 0}    & & \\
            & {\tt p}   & {\tt 5 -1 -4}  &             & & Declare $p_5 = \obar{x}_1 \pand \obar{p_4}$ \\
    {\tt 6} & {\tt ab}  & {\tt 5 1 4 0} & {\tt 0}    & & Defining clauses for $p_5$ \\
    {\tt 7} & {\tt ab}  & {\tt -5 -1 0} & {\tt -6 0}    & & \\
    {\tt 8} & {\tt ab}  & {\tt -5 -4 0} & {\tt -6 0}    & & \\
            & {\tt p}   & {\tt 6 1 2} &              & & Declare $p_6 = x_1 \pand x_2$ \\    
    {\tt 9} & {\tt ab}  & {\tt 6 -1 -2 0}  & {\tt 0}    & & Defining clauses for $p_6$ \\ 
    {\tt 10} & {\tt ab}  & {\tt -6 1 0}    & {\tt -9 0} & & \\
    {\tt 11} & {\tt ab}  & {\tt -6 2 0}    & {\tt -9 0} & & \\ 
            & {\tt s}   & {\tt 7 5 6}   & {\tt 7 10 0}  & & Declare $s_7 = p_5 \por p_6$ \\
    {\tt 12} & {\tt ab}  & {\tt -7 5 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
    {\tt 13} & {\tt ab}  & {\tt  7 -5 0}    & {\tt -12 0} & & \\  
    {\tt 14} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -12 0} & & \\

    {\tt 15} & {\tt ar}  & {\tt 1 -4 0}    & {\tt 4 5 1 0} & & Justify $\obar{x}_1 \rightarrow \obar{p_4}$ \\
    {\tt 16} & {\tt ar}  & {\tt 1 7 0}     & {\tt 13 6 15 0} & & Justify $\obar{x}_1 \rightarrow s_7$ \\
    {\tt 17} & {\tt ar}  & {\tt -1 7 0}    & {\tt 14 9 2 0} & & Justify $x_1 \rightarrow s_7$ \\
    {\tt 18} & {\tt ar}  & {\tt 7 0}       & {\tt 16 17 0}  & & Justify unit clause $t = s_7$ \\
             & {\tt dr}  & {\tt 1}         & {\tt 3 8 10 12 18 0} & & Delete $C_1$\\
             & {\tt dr}  & {\tt 2}         & {\tt 11 7 12 18 0} & & Delete $C_2$\\
  \end{tabular}
}  
  \caption{CRAT Proof \#2B for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$}
  \label{fig:p2-cdcl:crat}
\end{figure}

Figure~\ref{fig:p2-cdcl:schema} shows an alternate schema representing
the same formula $\phi_2$ as in Example~2A, and
Figure~\ref{fig:p2-cdcl:crat} shows the associated CRAT proof.  This
proof was generated via a top-down strategy, such as would be created
using a model counter based on CDCL\@.  It starts by splitting on
variable $x_1$.  Clause $C_1$ is trivially satisfied when $x_1$ is
assigned $\tautology$, and clause $C_2$ becomes the clause $C'_2 =
x_2$.  On the other hand, clause $C_2$ is trivially satisfied when
$x_1$ is assigned $\nil$, and clause $C_1$ becomes $C'_1 = x_2 \lor
x_3$.  Clause $C'_1$ can be represented schematically as
$\neg[\obar{x}_2 \pand \obar{x}_3]$, with this operation labeled $p_4$
in Figure~\ref{fig:p2-cdcl:schema}A\@.  The splitting on variable
$x_1$ can be rejoined as $\ite(x_1, x_2, \obar{p_4})$, and this ITE
operation can be expressed using the product operations $p_5$ and
$p_6$, joined by the sum operation $t = s_7$.

The two schemas shown in Figures~\ref{fig:p2-bdd:schema}A and
\ref{fig:p2-cdcl:schema}A have similar structure.  They have very
different negation patterns, but they are logically equivalent.  Their
associated cost computations (Figures~\ref{fig:p2-bdd:schema}B and
\ref{fig:p2-cdcl:schema}B) also yield the same results for arbitrary
weights $w_1$, $w_2$, and $w_3$.

The CRAT proof for the top-down approach follows a different pattern than does the proof based on a bottom-up construction.
It does
not create any intermediate unit clauses.  Instead, it constructs a proof
that the root $t = s_7$ holds as a unit clause by splitting into the
two assignments for variable $x_1$.  The proof that
$\obar{x}_1 \rightarrow s_7$ (proof line 16) builds on the proof that
$\obar{x}_1 \rightarrow \obar{p}_4$ (proof line 15), which derives from clause $C_1$.
The proof that $x_1 \rightarrow x_7$ (proof line 17) derives from clause $C_2$.
These two are combined to yield the unit clause $s_7$ (proof line 18).
Given this unit clause, the two input clauses can then be deleted.

@@

Given an arbitrary CNF formula, we can use BDD operations to generate
a schematic representation.  The proof generation can follow the
methods we have used for generating unsatisfiability proofs of Boolean
formulas~\cite{bryant:tacas:2021} and dual proofs of quantified
Boolean formulas~\cite{bryant:cade:2021}.
Each BDD node can be expressed as an ITE operation and
make use of the encodings shown in Table~\ref{tab:ite}.

A second class of model counters proceeds top-down, based on the CDCL
framework.  These choose a splitting variable $x$ and recursively
construct schemas $S_1$ and $S_0$ for the two assignments to the
variable.  These are combined as $\ite(x, S_1, S_0)$, using one of the
encodings of $\ite$ shown in Table~\ref{tab:ite}.  Like CDCL, the
top-down algorithm can make use of unit propagation, conflict
detection, and clause learning.  It can also make use of {\em
  variable} partitioning.  That is, suppose for some partial
assignment to the variables, the input clauses decompose into two or
more sets over disjoint variables.  Then the schema for each of these
partitions can be generated separately, and these are joined via the $\pand$
operation.
